---
title: "Machine Learning Datasets"
author: "Arnab Aich"
output:
  html_document:
    toc: true
    highlight: tango
    code_folding: hide
    fig_width: 8
    fig_height: 6
    fig_align: "center"
    self_contained: true
    toc_depth: 3
    toc_float:
      collapsed: true
---

<link rel="stylesheet" type="text/css" href="../styles.css">

# Abalone Dataset

The **Abalone dataset** is commonly used for regression tasks in machine learning. The main objective is to predict the age of abalone from its physical measurements. The age of an abalone is determined by counting the number of rings in its shell, which is approximately equal to the number of years the abalone has been alive (each ring represents about 1.5 years).

### Key Features:

- **Features**: The dataset includes eight attributes such as length, diameter, height, whole weight, shucked weight, viscera weight, shell weight, and sex (categorical: Male, Female, Infant).
- **Target**: The target variable is the number of **rings**, which is used to estimate the age of the abalone.

### Attributes:

- **Sex**: Categorical variable (M = Male, F = Female, I = Infant).
- **Length**: Longest shell measurement (in mm).
- **Diameter**: Diameter perpendicular to the length (in mm).
- **Height**: Height with meat in the shell (in mm).
- **Whole weight**: Whole abalone weight (in grams).
- **Shucked weight**: Weight of the meat (in grams).
- **Viscera weight**: Gut weight after bleeding (in grams).
- **Shell weight**: Weight of the shell after drying (in grams).
- **Rings**: Number of rings, which estimates the age (target variable).

The dataset contains **4,177 instances** of abalone records.

### How to Load the Abalone Dataset in R

The **Abalone dataset** is available in the UCI Machine Learning Repository. You can load the dataset directly into R using the following code:

```{r load-abalone-data}
# Load the Abalone dataset from UCI Machine Learning Repository

# URL where the dataset is stored
url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.data"
abalone <- read.csv(url, header = FALSE, stringsAsFactors = TRUE, sep = ",")
```

# Dexter Dataset

**Dexter** is a binary classification dataset from the NIPS 2003 feature selection challenge. It is a text classification task, where many features are irrelevant, making it a great dataset for testing feature selection algorithms.

### Key Features:
- **20,000 features**: Many of these features are sparse and irrelevant.
- **600 instances**: Samples in the dataset.
- **Task**: Binary classification of text documents.

### How to Download the Dexter Dataset

The Dexter dataset can be downloaded from the UCI Machine Learning Repository. Follow these steps to access and download the data:

1. **Go to the UCI Machine Learning Repository**:
   - Visit the [Dexter Dataset page on the UCI Repository](https://archive.ics.uci.edu/ml/datasets/Dexter).

2. **Download the following files from the Data Folder**:
   - `dexter_train.data`: Training data containing features.
   - `dexter_train.labels`: Labels for the training set.
   - `dexter_valid.data`: Validation data.
   - `dexter_valid.labels`: Labels for the validation set.
   - `dexter_test.data`: Test data (without labels).

3. **Unzip the files if necessary**:
   If the files are compressed, unzip them using a tool like `7-Zip` or your system's default extractor.

### Load the Dexter Dataset in R

After downloading the dataset, you can load it into R using the `read.table()` function. Below is an example of how to load and view the Dexter dataset in R.

```{r load-dexter, eval=FALSE}
# Load training data
dexter_train <- read.table("path_to_file/dexter_train.data", header = FALSE)

# Load training labels
dexter_labels <- read.table("path_to_file/dexter_train.labels", header = FALSE)

# View the structure of the training dataset
str(dexter_train)
```

# Gisette Dataset

**Gisette** is a binary classification dataset from the NIPS 2003 feature selection challenge. The task is to distinguish between handwritten digits '4' and '9'. The dataset contains a large number of features, many of which are noisy, making it suitable for feature selection research.

### Key Features:
- **5,000 features**: Includes both relevant and irrelevant (noisy) features.
- **13,500 instances**: Samples in the dataset.
- **Task**: Binary classification of handwritten digits '4' and '9'.

### How to Download the Gisette Dataset

The Gisette dataset can be downloaded from the UCI Machine Learning Repository. Follow these steps to access and download the data:

1. **Go to the UCI Machine Learning Repository**:
   - Visit the [Gisette Dataset page on the UCI Repository](https://archive.ics.uci.edu/ml/datasets/Gisette).

2. **Download the following files from the Data Folder**:
   - `gisette_train.data`: Training data containing features.
   - `gisette_train.labels`: Labels for the training set.
   - `gisette_valid.data`: Validation data.
   - `gisette_valid.labels`: Labels for the validation set.
   - `gisette_test.data`: Test data (without labels, typically for challenge purposes).

3. **Unzip the files if necessary**:
   If the files are compressed, unzip them using a tool like `7-Zip` or your system's default extractor.

### Load the Gisette Dataset in R

After downloading the dataset, you can load it into R using the `read.table()` function. Below is an example of how to load and view the Gisette dataset in R.

```{r load-gisette, eval=FALSE}
# Load training data
gisette_train <- read.table("path_to_file/gisette_train.data", header = FALSE)

# Load training labels
gisette_labels <- read.table("path_to_file/gisette_train.labels", header = FALSE)

# View the structure of the training dataset
str(gisette_train)
```


# Hill-Valley Dataset

The **Hill-Valley** dataset is a time-series classification dataset. The task is to determine whether a sequence of numbers represents a "hill" (upward trend followed by a downward trend) or a "valley" (downward trend followed by an upward trend). This dataset is useful for testing classification models in time-series data.

### Key Features:
- **100 features**: Continuous values in each sequence, representing the time series.
- **1,218 instances**: Total samples in the dataset.
- **2 classes**: Binary classification (hill or valley patterns).

### How to Download the Hill-Valley Dataset

The Hill-Valley dataset can be downloaded from the UCI Machine Learning Repository. Follow these steps to access and download the data:

1. **Go to the UCI Machine Learning Repository**:
   - Visit the [Hill-Valley Dataset page on the UCI Repository](https://archive.ics.uci.edu/ml/datasets/Hill-Valley).

2. **Download the following files from the Data Folder**:
   - `Hill_Valley_with_noise_Training.data`: Training data with noise.
   - `Hill_Valley_with_noise_Testing.data`: Test data with noise.
   - `Hill_Valley_without_noise_Training.data`: Training data without noise.
   - `Hill_Valley_without_noise_Testing.data`: Test data without noise.

3. **Unzip the files if necessary**:
   If the files are compressed, unzip them using a tool like `7-Zip` or your system's default extractor.

### Load the Hill-Valley Dataset in R

After downloading the dataset, you can load it into R using the `read.table()` function. Below is an example of how to load and view the Hill-Valley dataset in R.

```{r load-hill-valley, eval=FALSE}
# Load training data with noise
hill_valley_train <- read.table("path_to_file/Hill_Valley_with_noise_Training.data", header = FALSE)

# Load test data with noise
hill_valley_test <- read.table("path_to_file/Hill_Valley_with_noise_Testing.data", header = FALSE)

# View the structure of the training dataset
str(hill_valley_train)
```



# Madelon Dataset

The **Madelon dataset** is a synthetic dataset that was originally created for the NIPS 2003 feature selection challenge. It is widely used in machine learning research for benchmarking classification algorithms, particularly in feature selection and high-dimensional data analysis.

### Key Features:
- **500 features**: Many of these features are noise or irrelevant to the target classification.
- **Binary classification**: The target variable consists of two balanced classes (1 and -1).
- **2600 samples**: It includes 2000 training samples and 600 validation samples.

The dataset is challenging because individual features have little predictive power, but complex interactions between them can yield meaningful insights for classification tasks. This makes Madelon ideal for testing feature selection algorithms or classification methods that can capture interactions between features.

### How to Download the Madelon Dataset

The Madelon dataset can be downloaded from the UCI Machine Learning Repository. Follow these steps to access and download the data:

1. **Go to the UCI Machine Learning Repository:**
   - Visit the [Madelon Dataset page on the UCI Repository](https://archive.ics.uci.edu/ml/datasets/Madelon).

2. **Download the following files from the Data Folder**:
   - `madelon_train.data`: The training set with 500 features for each sample.
   - `madelon_train.labels`: Labels for the training set.
   - `madelon_valid.data`: The validation set with 500 features for each sample.
   - `madelon_valid.labels`: Labels for the validation set.

3. **Unzip the files if necessary**:
   If any of the files are compressed, extract them using appropriate software like `7-Zip` or the built-in extraction tool in your operating system.

4. **Load the data into R**:
   After downloading the data, you can load it into R using the `read.table()` function.



# Satimage Dataset

The **Satimage dataset** comes from the Landsat satellite imagery and is used for classification tasks. The dataset contains multi-spectral values of pixels from satellite images, and the task is to classify these pixels into one of six different types of soil or vegetation. This dataset is commonly used in remote sensing and pattern recognition applications.

### Key Features:
- **36 features**: Each feature represents a spectral band from satellite images.
- **6 classes**: The dataset consists of six different classes representing different types of soil and vegetation.
- **6435 samples**: The dataset has 4435 training samples and 2000 test samples.

The Satimage dataset is challenging because it involves multi-dimensional data that captures pixel information from satellite images. It is often used to benchmark classification algorithms in remote sensing and image analysis.

### How to Download the Satimage Dataset

The Satimage dataset can be downloaded from the UCI Machine Learning Repository. Follow these steps to access and download the data:

1. **Go to the UCI Machine Learning Repository**:
   - Visit the [Satimage Dataset page on the UCI Repository](https://archive.ics.uci.edu/ml/datasets/Statlog+(Landsat+Satellite)).

2. **Download the following files from the Data Folder**:
   - `sat.trn`: Training data containing multi-spectral pixel values.
   - `sat.tst`: Test data for evaluating the model.
   - `sat.train`: Combined training and test data in one file (optional).
   - `sat.tstcls`: True class labels for the test data.

3. **Unzip the files if necessary**:
   If the files are compressed, unzip them using a tool like `7-Zip` or your system's default extractor.

4. **Load the data into R**:
   After downloading the data, you can load it into R using the `read.table()` function.



